# The impact of Storms and other severe weather events on public health and the economy of the United States

## Synopsis

This report contains a brief digest of the 60-year spanning US weather incidents data base. We present results with regards to two investigations: The most severe events in terms of public health and safety and the most severe events in terms of total economic impact. Tornados, extreme heat and floods turn out to be the deadliest and most dangerous weather events. The events with the largest economic impact are floods, hurricanes and tornados.

## Data Processing
Loading useful libraries:
```{r cache=TRUE}
library("stringr")
```
Loading the data and caching it:
```{r cache=TRUE}
#following needs to be cached

classes = c("character", "character", "NULL", "NULL", "NULL", "NULL", "character", "character", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "character", "character", "character", "character", "character", "character", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "numeric")

st <- read.csv("repdata-data-StormData.csv.bz2", header = TRUE, colClasses = classes, nrows=902297,comment.char="")

# I dont think I need the following:
st$BGN_DATE <- NULL

st$STATE__ <- NULL
st$STATE <- NULL
```
Fixing some of the events data character strings:
```{r cache=TRUE}
st$EVTYPE <- tolower(st$EVTYPE)
st$EVTYPE <- str_trim(st$EVTYPE)
st$EVTYPE <- sub("\\\\",".",st$EVTYPE)
st$EVTYPE <- sub("precipitation","rain",st$EVTYPE)
st$EVTYPE <- sub("avalance","avalanche",st$EVTYPE)
st$EVTYPE <- sub("dense fog","fog",st$EVTYPE)
st$EVTYPE <- sub("freezing fog","fog",st$EVTYPE)
st$EVTYPE <- sub("heavy surf/high surf","heavy surf",st$EVTYPE)
#lightning
st$EVTYPE <- sub("lightning injury","lightning",st$EVTYPE)
st$EVTYPE <- sub("lightning fire","lightning",st$EVTYPE)
#heat
st$EVTYPE <- sub("excessive heat","heat",st$EVTYPE)
st$EVTYPE <- sub("unseasonably warm","heat",st$EVTYPE)
st$EVTYPE <- sub("heat wave drought","heat",st$EVTYPE)
st$EVTYPE <- sub("heat and dry","heat",st$EVTYPE)
st$EVTYPE <- sub("record heat","heat",st$EVTYPE)
st$EVTYPE <- sub("extreme heat","heat",st$EVTYPE)
st$EVTYPE <- sub("heat waves","heat",st$EVTYPE)
st$EVTYPE <- sub("heat wave","heat",st$EVTYPE)
st$EVTYPE <- sub("heats","heat",st$EVTYPE)
#hurricane
st$EVTYPE <- sub("hurricane edouard","hurricane",st$EVTYPE)
st$EVTYPE <- sub("hurricane erin","hurricane",st$EVTYPE)
st$EVTYPE <- sub("hurricane opal/high winds","hurricane",st$EVTYPE)
st$EVTYPE <- sub("hurricane/typhoon","hurricane",st$EVTYPE)
st$EVTYPE <- sub("hurricane emily","hurricane",st$EVTYPE)
st$EVTYPE <- sub("hurricane opal","hurricane",st$EVTYPE)
st$EVTYPE <- sub("hurricane gordon","hurricane",st$EVTYPE)
st$EVTYPE <- sub("hurricane felix","hurricane",st$EVTYPE)
st$EVTYPE <- sub("hurricane-generated swells","hurricane",st$EVTYPE)
#thunderstorm
st$EVTYPE <- sub("tstm wind (g40)","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("tstm wind (g45)","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thundersnow","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("severe thunderstorm","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstorm  winds","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstorm winds","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstorm winds/hail","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstorms winds","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstorm wind","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstorm winds 13","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstorm windss","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstorm 13","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstorms","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("thunderstormw","thunderstorm",st$EVTYPE)
st$EVTYPE <- sub("tstm wind","thunderstorm",st$EVTYPE)
#storm
st$EVTYPE <- sub("winter storm high winds","storm",st$EVTYPE)
st$EVTYPE <- sub("winter storms","storm",st$EVTYPE)
st$EVTYPE <- sub("tropical storm gordon","storm",st$EVTYPE)
st$EVTYPE <- sub("tropical storm","storm",st$EVTYPE)
st$EVTYPE <- sub("winter storm","storm",st$EVTYPE)
st$EVTYPE <- sub("storm dean","storm",st$EVTYPE)
st$EVTYPE <- sub("storm alberto","storm",st$EVTYPE)
st$EVTYPE <- sub("storm jerry","storm",st$EVTYPE)
st$EVTYPE <- sub("high wind 48","storm",st$EVTYPE)
st$EVTYPE <- sub("storm surge/tide","storm surge",st$EVTYPE)
```
Continuing with events name adjustment:
```{r cache=TRUE}
#wind
st$EVTYPE <- sub("high winds","wind",st$EVTYPE)
st$EVTYPE <- sub("high wind and seas","wind",st$EVTYPE)
st$EVTYPE <- sub("high wind/seas","wind",st$EVTYPE)
st$EVTYPE <- sub("high winds/cold","wind",st$EVTYPE)
st$EVTYPE <- sub("winds","wind",st$EVTYPE)
st$EVTYPE <- sub("strong wind","wind",st$EVTYPE)
st$EVTYPE <- sub("high wind","wind",st$EVTYPE)
#wind/snow
st$EVTYPE <- sub("high wind/heavy snow","wind/snow",st$EVTYPE)
st$EVTYPE <- sub("high winds/snow","wind/snow",st$EVTYPE)
st$EVTYPE <- sub("snow/wind","wind/snow",st$EVTYPE)
st$EVTYPE <- sub("heavy snow and wind","wind/snow",st$EVTYPE)
#cold
st$EVTYPE <- sub("cold wave","cold",st$EVTYPE)
st$EVTYPE <- sub("extreme cold","cold",st$EVTYPE)
st$EVTYPE <- sub("extended cold","cold",st$EVTYPE)
st$EVTYPE <- sub("cold weather","cold",st$EVTYPE)
st$EVTYPE <- sub("record cold","cold",st$EVTYPE)
st$EVTYPE <- sub("ccold/wind chill","cold",st$EVTYPE)
st$EVTYPE <- sub("unseasonably cold","cold",st$EVTYPE)
#ice
st$EVTYPE <- sub("icy roads","ice",st$EVTYPE)
st$EVTYPE <- sub("ice roads","ice",st$EVTYPE)
st$EVTYPE <- sub("ice jam flooding","ice",st$EVTYPE)
#flood
st$EVTYPE <- sub("coastal  flood/erosion","flood",st$EVTYPE)
st$EVTYPE <- sub("coastal  flooding/erosion","flood",st$EVTYPE)
st$EVTYPE <- sub("coastal flooding/erosion","flood",st$EVTYPE)
st$EVTYPE <- sub("coastal flood","flood",st$EVTYPE)
st$EVTYPE <- sub("rain and flood","flood",st$EVTYPE)
st$EVTYPE <- sub("flash flooding/flood","flood",st$EVTYPE)
st$EVTYPE <- sub("flash flood/flood","flood",st$EVTYPE)
st$EVTYPE <- sub("erosion/cstl flood","flood",st$EVTYPE)
st$EVTYPE <- sub("coastal surge","flood",st$EVTYPE)
st$EVTYPE <- sub("river and stream flood","flood",st$EVTYPE)
st$EVTYPE <- sub("major flood","flood",st$EVTYPE)
st$EVTYPE <- sub("flood/flash flood","flood",st$EVTYPE)
st$EVTYPE <- sub("river flooding","flood",st$EVTYPE)
st$EVTYPE <- sub("river flood","flood",st$EVTYPE)
st$EVTYPE <- sub("flash flooding","flood",st$EVTYPE)
st$EVTYPE <- sub("flood/flood","flood",st$EVTYPE)
st$EVTYPE <- sub("flash flood/","flood",st$EVTYPE)
st$EVTYPE <- sub("flash flood","flood",st$EVTYPE)
st$EVTYPE <- sub("flood/flash","flood",st$EVTYPE)
st$EVTYPE <- sub("urban/sml stream fld","flood",st$EVTYPE)
st$EVTYPE <- sub("urban flood","flood",st$EVTYPE)
st$EVTYPE <- sub("heavy rains/flooding","flood",st$EVTYPE)
st$EVTYPE <- sub("heavy rain/flooding","flood",st$EVTYPE)
st$EVTYPE <- sub("snowmelt flooding","flood",st$EVTYPE)
st$EVTYPE <- sub("flood & heavy rain","flood",st$EVTYPE)
st$EVTYPE <- sub("flooding","flood",st$EVTYPE)
st$EVTYPE <- sub("floods","flood",st$EVTYPE)
#winter
st$EVTYPE <- sub("winter weather mix","winter",st$EVTYPE)
st$EVTYPE <- sub("wintry mix","winter",st$EVTYPE)
st$EVTYPE <- sub("winter weather","winter",st$EVTYPE)
st$EVTYPE <- sub("winter weather/mix","winter",st$EVTYPE)
st$EVTYPE <- sub("winter/mix","winter",st$EVTYPE)
#tornado
st$EVTYPE <- sub("tornado f2","tornado",st$EVTYPE)
st$EVTYPE <- sub("tornadoes, tstm wind, hail","tornado",st$EVTYPE)
st$EVTYPE <- sub("tornado f3","tornado",st$EVTYPE)
st$EVTYPE <- sub("tornado f1","tornado",st$EVTYPE)
#rip current
st$EVTYPE <- sub("rip currents/heavy surf","rip current",st$EVTYPE)
st$EVTYPE <- sub("rip currents","rip current",st$EVTYPE)
#rain
st$EVTYPE <- sub("heavy rains","rain",st$EVTYPE)
st$EVTYPE <- sub("excessive rainfall","rain",st$EVTYPE)
st$EVTYPE <- sub("freezing rain","rain",st$EVTYPE)
st$EVTYPE <- sub("heavy rain","rain",st$EVTYPE)
st$EVTYPE <- sub("rain/severe weather","rain",st$EVTYPE)
st$EVTYPE <- sub("freezing drizzle","rain",st$EVTYPE)
#snow
st$EVTYPE <- sub("heavy snow","snow",st$EVTYPE)
st$EVTYPE <- sub("light snow","snow",st$EVTYPE)
st$EVTYPE <- sub("snow squall","snow",st$EVTYPE)
st$EVTYPE <- sub("record snow","snow",st$EVTYPE)
st$EVTYPE <- sub("excessive snow","snow",st$EVTYPE)
st$EVTYPE <- sub("snow/snow","snow",st$EVTYPE)
#wildfire
st$EVTYPE <- sub("wild/forest fire","wildfire",st$EVTYPE)
st$EVTYPE <- sub("wild fires","wildfire",st$EVTYPE)
#frost
st$EVTYPE <- sub("hard freeze","frost",st$EVTYPE)
st$EVTYPE <- sub("damaging freeze","frost",st$EVTYPE)
st$EVTYPE <- sub("frost/freeze","frost",st$EVTYPE)
st$EVTYPE <- sub("agricultural freeze","frost",st$EVTYPE)
st$EVTYPE <- sub("early frost","frost",st$EVTYPE)
st$EVTYPE <- sub("frost.freeze","frost",st$EVTYPE)
st$EVTYPE <- sub("freeze","frost",st$EVTYPE)
#hail
st$EVTYPE <- sub("small hail","hail",st$EVTYPE)
```
Processing the damage data, converting all entries to numerics (since they symbolise are exponents):
```{r cache=TRUE}

st$CROPDMGEXP[st$CROPDMGEXP== ""] <- "0"
st$PROPDMGEXP[st$PROPDMGEXP==""] <- "0"

st$CROPDMGEXP[st$CROPDMGEXP=="K"] <- "3"
st$CROPDMGEXP[st$CROPDMGEXP=="k"] <- "3"

st$PROPDMGEXP[st$PROPDMGEXP=="K"] <- "3"

st$CROPDMGEXP[st$CROPDMGEXP=="M"] <- "6"
st$PROPDMGEXP[st$PROPDMGEXP=="M"] <- "6"

st$CROPDMGEXP[st$CROPDMGEXP=="m"] <- "6"
st$PROPDMGEXP[st$PROPDMGEXP=="m"] <- "6"

st$CROPDMGEXP[st$CROPDMGEXP=="H"] <- "2"
st$PROPDMGEXP[st$PROPDMGEXP=="H"] <- "2"

st$CROPDMGEXP[st$CROPDMGEXP=="?"] <- "0"

st$PROPDMGEXP[st$PROPDMGEXP=="h"] <- "2"

st$CROPDMGEXP[st$CROPDMGEXP=="B"] <- "9"
st$PROPDMGEXP[st$PROPDMGEXP=="B"] <- "9"

st$PROPDMGEXP[st$PROPDMGEXP=="?"] <- "0"
st$PROPDMGEXP[st$PROPDMGEXP=="+"] <- "0"
st$PROPDMGEXP[st$PROPDMGEXP=="-"] <- "0"
```
We end up with the new more tidy exponential terms which we then use to create one new collumn with the actual total amount in millions of USD:
```{r cache=TRUE}
st$USD <- apply(st,1, function(row) 0.000001*(as.integer(row[6])*10^(as.integer(row[7])) + as.integer(row[4])*10^(as.integer(row[5]))))

st$PROPDMGEXP <- NULL
st$CROPDMGEXP <- NULL
st$PROPDMG <- NULL
st$CROPDMG <- NULL
```
Processing the fatalities and injuries data:
```{r cache=FALSE}
st$FATALITIES <- as.integer(st$FATALITIES)
st$INJURIES <- as.integer(st$INJURIES)
```

Selecting the data with the most fatalities and injuries, disregarding the low-casualty/injury ones:

```{r cache=TRUE}
st_health <- st[!(st$INJURIES < 10 & st$FATALITIES < 10),]
```
Analysing events with health impact:
```{r cache=TRUE}
st_health$EVTYPE <- as.factor(st_health$EVTYPE)

total_casualties <- aggregate(st_health$FATALITIES ~ st_health$EVTYPE, FUN=sum)
reorder_casualties <- total_casualties[order(-total_casualties[,2]),]
names(reorder_casualties) <- c("Event Type", "Number of Casualties")

total_injured <- aggregate(st_health$INJURIES ~ st_health$EVTYPE, FUN=sum)

reorder_injured <- total_injured[order(-total_injured[,2]),]
names(reorder_injured) <- c("Event Type", "Number of Injuries")
```
Processing the data for economic impact in a similar manner:
```{r cache=TRUE}
st_economy <- st[!(st$USD < 0.08),]
```
Analysing events with economic impact:
```{r}
st_economy$EVTYPE <- tolower(st_economy$EVTYPE)
st_economy$EVTYPE <- as.factor(st_economy$EVTYPE)

total_economy <- aggregate(st_economy$USD ~ st_economy$EVTYPE, FUN=sum)
reorder_economy <- total_economy[order(-total_economy[,2]),]
names(reorder_economy) <- c("Event Type", "Total Cost in millions USD")
```
## Results

Top-5 most harmful weather events with respect to population health: Injuries
```{r results = 'asis'}
kable(reorder_injured[1:5,], format="markdown", row.names=F)

barplot(as.vector(reorder_injured[1:5,2]), names.arg=as.vector(reorder_injured[1:5,1]), main="The five weather events that cause the most injuries")
```
Top-5 most harmful weather events with respect to population health: Casualties
```{r results = 'asis'}
kable(reorder_casualties[1:5,], format="markdown", row.names=F)

barplot(as.vector(reorder_casualties[1:5,2]), names.arg=as.vector(reorder_casualties[1:5,1]), main="The five most fatal weather events in the US")
```
Top-5 weather events with the greatest economic consequences:
```{r results = 'asis'}
kable(reorder_economy[1:5,], format="markdown", row.names=F)

barplot(as.vector(reorder_economy[1:5,2]), names.arg=as.vector(reorder_economy[1:5,1]), main="The five most costly weather events in the US in million USD")
```
## Recommendations to the US Government

Divest from fossil fluels ASAP. You are welcome.